---
title: "Creating Bar Charts and Map Visualizations in R"
author: "Jottin Lugo"
date: "July 1, 2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```
## Introduction
I created a simple data analysis report using `knitr` from Rstudio. The report consists of bar charts generated with `ggplot()` and map visualization from `plot_usmap()` and `ggplotly()`. There are several other libraries that I used to complete this task which are listed on the section below.

#### Loading the libraries
```{r libraries, echo=TRUE}
# prior to loading libraries you must install them
# install.packages("libraryname")
library(janitor) # easily reformat column names using the clean_names function
library(dplyr) # sorts the data using the arrange and desc function
library(tidyr) # slice the data using the slice function
library(ggplot2) # plots the data
library(stringr) # truncate axis titles when needed
library(scales) # convert values to dollar format
library(usmap) # generates a static map
library(plotly) # makes map interactive by handling text hovering the map
library(readr) # handles the import of csv file: states_longitude_latitude.csv
```

#### Filtering and Sorting the Data
```{r clean, echo=TRUE}
# separate the data in two groups: income and rent, then assign this to a variable
rent_data <- clean_names(us_rent_income) %>% filter(variable=="rent") %>% arrange(desc(estimate))
income_data <- clean_names(us_rent_income) %>% filter(variable=="income") %>% arrange(desc(estimate))
```

#### Organizing the Data
```{r slice, echo=TRUE}
# returns the top 10 results, ordered by specific column, then assign this to a variable
top_10_rent <- slice_max(rent_data, order_by = estimate, n = 10)
top_10_income <- slice_max(income_data, order_by = estimate, n = 10)
```

#### Variables
```{r variables, echo=TRUE}
# values that repeat are saved in a variable to clear up space
source <- "Source: tidyr - us_rent_income data"
```

#### Getting the Map Data Ready for Visualization
```{r mapdata, echo=TRUE}
# read csv file located in working directory
states_coord <- read_csv("states_longitude_latitude.csv")

# rename column name back to state before plotting
states_coord_as_name <- rename(states_coord, name=state)

# join data frames by column name
joined_data_income_final <- inner_join(income_data,states_coord_as_name,by="name")

# convert to data frame and rename the column back to state
income_final <- data.frame(joined_data_income_final) %>% rename(state=name)

```

## Summary
Now that I've loaded the libraries, filtered, sorted, and sliced the data I am going to use in this project; I can plot it using *ggplot2*. For this example, the function `geom_col()` creates the bar charts. To zoom in on the data, I used the `coord_cartesian()` function.

#### Figure 1
```{r income, echo=TRUE, out.width="90%"}
# chart 1 ggplot and geom_col generate the bar charts
ggplot(top_10_income,
       mapping = aes(
         x = str_trunc(name, 20), # str_trunc function makes sure names do not overlap
         y = estimate,
         fill = estimate
       )) +
  geom_col() +
  geom_text(aes(label = dollar(estimate)),
            size = 3,
            color = "white",
            vjust = 1.5) +
  scale_y_continuous(labels = dollar) +
  coord_cartesian(xlim = c(1, NA), ylim = c(30000, NA)) + # coord_cartesian function Zooms in Y-Axis and scales view
  labs(
    title = "Top 10 Highest Income in the United States",
    x = "State",
    y = "Income",
    caption = source,
    fill = "Income"
  ) +
  theme_light(base_size = 9) + # theme_light to handle most of aesthetics
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14),
    plot.caption = element_text(color = "#5C6380")
  )
```

#### Figure 2
```{r rent, echo=TRUE, out.width="90%"}
# chart 2 ggplot and geom_col generate the bar charts
ggplot(top_10_rent,
       mapping = aes(
         x = str_trunc(name, 20),
         y = estimate,
         fill = estimate
       )) + 
  geom_col() +
  geom_text(aes(label = dollar(estimate)),
            size = 3,
            color = "white",
            vjust = 1.5) +
  scale_y_continuous(labels = dollar) +
  coord_cartesian(xlim = c(1, NA), ylim = c(1000, NA)) +
  labs(
    title = "Top 10 Highest Rent in the United States",
    x = "State",
    y = "Rent",
    caption = source,
    fill = "Rent"
  ) +
  theme_light(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14),
    plot.caption = element_text(color = "#5C6380")
  )
```

### Preparing the Data For Questions
```{r prepare, echo=TRUE}
income_sum <- drop_na(income_final) %>% summarize(sum(estimate))
rent_sum <- drop_na(rent_data) %>% summarize(sum(estimate))
perc_income_to_rent <- round(income_sum$`sum(estimate)` / rent_sum$`sum(estimate)`, 2)
top_10_lowest_income <- head(joined_data_income_final %>% arrange(estimate), n = 10)
```

### What Percentage of the income is going to rent?
```{r insight1, echo=TRUE}
print(paste(perc_income_to_rent,"%"))
```

### What's the min, median, and max value of income?
```{r insight2, echo=TRUE}
income_final$estimate %>% summary()
```
 
### What's the min, median, and max value of rent?
```{r insight3, echo=TRUE}
rent_data$estimate %>% summary()
```
 
### What are the top 10 states with lowest income?
```{r insight4, echo=TRUE}
top_10_lowest_income %>% select(name, estimate)
```


For the interactive map, I used `plot_usmap()` and `ggplotly()` from *plotly*: to generate the map visualization *plot_usmap()* and to act as interactive layer I used `ggplotly()`. The csv data was loaded with `readr()`. Additionally, I joined two data frames with `inner_join()` from the *dplyr* package.

#### Figure 3:
```{r mapview, echo=TRUE, out.width="100%"}
# map visualization made with usmap, plotly and ggplot2
p <- plot_usmap(
  data = data.frame(income_final),
  values = "estimate",
  color = "grey",
  labels = TRUE
  ) +
  scale_fill_continuous(name = "Income", labels = dollar) +
  theme_light() +
  theme(axis.text = element_blank()) + # hide the axis text for minimal viewing
  theme(legend.position = "right") +
  labs(title = "Income by State", y = "Latitude", x = "Longitude")
ggplotly(p)
```

## Closing
This is just one way to generate map visualizations in R. There are other libraries such as *rbokeh* or *leaflet* that allow you to create interactive maps. I was inclined to use *plotly* and *usmap* because they work well with *ggplot2* `theme()`. Further, `ggplotly()` and `plot_usmap()` help you to create swiftly USA interactive maps and that was part of the focus of this project.


For additional documentation, you can visit [usmap](https://cran.r-project.org/web/packages/usmap/vignettes/mapping.html), [plotly](https://plotly.com/r) and [ggplot2](https://ggplot2.tidyverse.org).

